---
title: "hw_03.Rmd"
author: "Vinay K L"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Q1

Suppose we have a dataset A (see code below) where each column represents multiple measures of nitrogen concentration in a particular lake. We want to get the average value for each lake. Do this in two ways: a for loop and a vectorized function colMeans().

set.seed(12) # to be reproducible
A = matrix(data = runif(n = 1:500), nrow = 50, ncol = 10)
colnames(A) = paste("lake", 1:10, sep = "_")

## Solution

```{r}
set.seed(12) # to be reproducible
A = matrix(data = runif(n = 1:500), nrow = 50, ncol = 10)
colnames(A) = paste("lake", 1:10, sep = "_")
```

Using for loop

```{r}
num_lakes <- ncol(A)
average_values1 <- numeric(num_lakes)

for (i in 1:num_lakes) {
  average_values1[i] <- mean(A[, i])
}

print(average_values1)
```

Using colMeans()

```{r}
average_values2 <- colMeans(A)

print(average_values2)
```


