---
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document: default
  html_document: default
link-citations: yes
bibliography: [references_final.bib, packages.bib]
indent: true
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
```

```{r, include=FALSE}
knitr::write_bib(c(.packages(), "ggplot2", "dplyr", "ggthemes", "stringr"), "packages.bib")
```

# Mapping the Void: An In-depth Analysis of the VertNet Database to Unearth Global Gaps in Avian Specimen and Tissue Collections

```{=tex}
\begin{center}
Vinay K. L.
\end{center}
```

## **Introduction**

Museums and natural history collections are important and invaluable sources of Earth history's biodiversity information. Often spanning centuries, museum collections represent a treasure trove of
biological specimens and data. The specimens in natural history
collections serve as the basis for numerous basic science and
research fields. Natural history collections also contain deep, taxon-specific
information derived from global populations of plants, animals, fungi,
and microorganisms. [@hope2018collection; @johnson2011climate;
@card2021museum] By evaluating how species evolve, museum
specimens offer crucial baseline data for studies of conservation and
emerging diseases (such as
[Hantavirus](https://msb.unm.edu/divisions/mammals/research/projects/hantavirus.html)
and West Nile Virus). The importance of museum collections lies in their
ability to serve as both a historical record of life on our planet and a
contemporary resource for scientific inquiry.

What initially began as cabinets filled with mounted specimens has
transformed into vast, digitized repositories. With the advent of modern
databases and digitization efforts, these collections are now more
accessible and interconnected than ever. VertNet is one such effort to
combine museum collection records from over 250 natural history
collections. [@constable2010vertnet] VertNet, a comprehensive repository
of biodiversity data, serves as an invaluable resource for assessing the
state of biodiversity worldwide. Exploiting VertNet's expansive
dataset facilitates a profound comprehension of the immense
diversity encompassed within avian taxa and provides the tools to
identify significant gaps in our intellectual understanding. By meticulously 
analyzing this wealth of information, we can discern
regions characterized by a need for avian tissue collections. These taxonomic
groups remain underrepresented in scientific investigations and
temporal gaps. These conclusions play a crucial part in determining how
future specimen collection initiatives will proceed. With this background, 
we set out to understand a) Where are the significant data gaps regarding specimen 
collection and tissue samples? 2) Which are the over and under-represented 
groups/families of birds in the museum collection. 

## **Methods**
The compilation of museum collection data was executed through two distinct 
methodologies. Firstly, data acquisition transpired through direct downloads 
from the [VertNet website](https://datacommons.cyverse.org/browse/iplant/home/shared/commons_repo/curated/Vertnet_Aves_Sep2016) and leveraged the rvertnet package,[@rvertnet] explicitly 
employing the 'bigsearch' option. Post retrieval, a meticulous sequence of data 
refinement procedures was implemented. The initial step involved merging  
two datasets, retaining only singular, distinct records. Notably, the VertNet 
dataset incorporated a spectrum of entries, including fossil specimens, eggs, 
and nest records, which were systematically removed from the dataset. Scrutiny 
of the unique country records exposed over 1100 entries, underscoring multiple 
nomenclatures for identical countries. Consequently, a consolidation process 
is employed, integrating countries and appending iso3c country codes by utilizing the countrycode package.[@countrycode] Subsequently, collection years were 
binned into 5-year intervals. 

Furthermore, a comprehensive data cleansing regimen was implemented to eliminate
any instances of missing genus and species names. Subsequently, orders and 
families underwent systematic refinement to standardize records in accordance 
with the IOC taxonomy.[@IOC_World_Bird_List] Upon completion of requisite columnar cleansing, the 
refined dataset was exported in CSV format, exclusively retaining selected 
columns for subsequent analytical procedures. 

The refined dataset later served as the foundational substrate for visualizing 
specimen and tissue collections spatially and temporally. The rnaturalearth 
package [@rnaturalearth] facilitated the importation of world map data, 
enabling the visualization of bird specimens and tissue collections across 
various collection years. 
Following the visualization process, a meticulous analysis ensued, wherein 
standardized residuals were calculated to find the significantly underrepresented 
and/or overrepresented groups of birds among order and family of birds. 
This methodological cascade underscores the rigor and precision 
applied to both the preprocessing and analytical phases of the investigation, 
enhancing the scholarly integrity and relevance of the ensuing findings. 

```{r, include=FALSE}
# This step import, cleans (Multi level) and exports the cleaned data set. 
#source("code/01_clean_data.R", local = knitr::knit_global())
```


```{r, include=FALSE}
df <- read.csv("data/cleaned_vertnet.csv")
```

## **Results**

The merging of two datasets culminated in a dataset comprising 1.508 million 
raw records of museum collections. Following meticulous data cleansing procedures, 
the resultant dataset retained approximately 1.18 million records, with approximately 
138,000 featuring tissue records. As illustrated in Figure 1, museum collections 
commencing from 1690 and extending to 2015, the collection data exhibits a notable 
acceleration from 1870 onwards, reaching its peak between 1928 and 1932 with an 
approximate tally of 23,000 specimens globally. 

```{r, fig.cap="Avian Specimen collection in museums across years binned to 5 years winddow"}
library(ggplot2)
library(ggthemes)
df <- read.csv("data/cleaned_vertnet.csv")

ggplot(na.omit(df), aes(x = as.factor(YearBin))) +
  geom_bar(fill = "#11A797") +
  labs(x = "Year of Collection", y = "No of Specimens") +
  ggtitle("Temporal distribution of avian specimen collection") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5)) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.5))
#ggsave("figs/Figure_1.jpg", plot = F1, width = 10, height = 7, units = "in")
```

Among the 237 countries catalogued, records are distributed across 190 nations, 
with conspicuous concentrations in the United States of America, as depicted in 
Figure 2. In contrast, the central region of Africa emerges as an undersampled 
center devoid of recorded specimens. The taxonomic exploration of class Aves, 
delineated by the IOC Bird taxonomy, elucidates a taxonomy comprising 41 orders 
and 255 families. The collection records, spanning 40 orders and 
232 families underscore a comprehensive representation of avian biodiversity. 

```{r, fig.cap="Spatial distribution of the avian specimen collections across globe. Red indicates high and yellow indicates low numbers whereas light grey shows the absence of specimen in logarithmic scale"}
library(ggthemes)
library(cowplot)
library(rnaturalearth)
library(dplyr)


count_data <- df %>%
  filter(!is.na(countrycode)) %>%  # Remove rows where countrycode is NA
  group_by(countrycode) %>%
  summarise(Count = n(), .groups = "drop")

world <- ne_countries(scale = "medium", returnclass = "sf", type = "countries")

merged_map_data <- merge(world, count_data, by.x = "iso_a3", by.y = "countrycode", all.x = TRUE)

ggplot() +
  geom_sf(data = merged_map_data, color = "black", fill = "lightgray") +
  geom_sf(data = merged_map_data, aes(fill = Count), color = NA) +
  scale_fill_gradient(low = "#F6EEAB", high = "#FF2400", na.value = "lightgray", guide = "legend", trans = "log")+
  ggtitle("Spatial distribution of avian specimen collection") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
#ggsave("figs/Figure_2.jpg", plot = F2, width = 10, height = 7, units = "in")

```

Coming to the tissue collection, a pronounced surge in "modern" tissue collection is discernible from the late 1970s, as depicted in Figure 3. 

```{r, fig.cap="Avian tissue only collection in museums across years binned to 5 years winddow"}
library(ggplot2)

tissue_df <- df[df$hastissue == 1, ]

ggplot(na.omit(tissue_df), aes(x = as.factor(YearBin))) +
  geom_bar(fill = "#11A797") +
  labs(x = "Year of Collection", y = "No of Specimens") +
  ggtitle("Temporal distribution of tissue collection") +
  theme_minimal() +
  #cale_colour_Publication() +
  theme(plot.title = element_text(hjust = 0.5)) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.5))
#ggsave("figs/Figure_3.jpg", plot = F3, width = 10, height = 7, units = "in")

```

The United States of America assumes a prominent role in tissue collection, 
contributing slightly over 60,000 specimens. In contrast to the broader specimen
collections, tissue collection encompasses only 145 countries, revealing a 
pronounced void in global representation. The taxonomic spectrum of tissue 
collection spans 37 orders and 197 families of avian taxa, accentuating the 
specificity and selectivity inherent in this facet of scientific 
exploration.  

```{r, fig.cap="Spatial distribution of the avian tissue collections across globe. Red indicates high and yellow indicates low numbers whereas light grey shows the absence of specimen in logarithmic scale"}
library(ggplot2)

count_data_tissue2 <- tissue_df %>%
  filter(!is.na(countrycode)) %>%  # Remove rows where countrycode is NA
  group_by(countrycode) %>%
  summarise(Count = n(), .groups = "drop")

world <- ne_countries(scale = "medium", returnclass = "sf")

merged_map_data_tissue <- merge(world, count_data_tissue2, by.x = "iso_a3", by.y = "countrycode", all.x = TRUE)

ggplot(na.omit(tissue_df)) +
  geom_sf(data = merged_map_data_tissue, color = "black", fill = "lightgray") +
  geom_sf(data = merged_map_data_tissue, aes(fill = Count), color = NA) +
  scale_fill_gradient(low = "#F6EEAB", high = "#FF2400", na.value = "lightgray", guide = "legend", trans = "log") +
  ggtitle("Spatial distribution of tissue collection") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

# ggsave("figs/Figure_4.jpg", plot = F4, width = 10, height = 7, units = "in")

```

```{r, fig.pos='H', fig.cap="Standardized Residuals - Orders. Chi-sqaure test p-value < 2.2e-16. Positive value indicates the overrepresentation and vice versa", include=TRUE}
# This is performing the Chi-square test. 
order_table <- table(df$order) # creating a table
order_chi_square <- chisq.test(order_table) # performing chi-square test

order_residuals <- resid(order_chi_square) # running residuals to check the representation

par(mar = c(9, 6, 4, 2) + 0.1)
barplot(order_residuals, names.arg = levels(df$Order), main = "Standardized Residuals - Orders",col = "#11A797", las = 2) # plotting the results
```

```{r, fig.pos='H', fig.cap="Standardized Residuals - Families. Chi-sqaure test p-value < 2.2e-16. Positive value indicates the overrepresentation and vice versa", include=TRUE}
family_table <- table(df$family) # creating a table
family_chi_square <- chisq.test(family_table) # perfomning Chi-square test
 # checking the results. 

family_residuals <- resid(family_chi_square) # running resuduals to check the represntation. 
 # checking the results

par(mar = c(9, 6, 4, 2) + 0.1)
barplot(family_residuals, names.arg = levels(df$family), main = "Standardized Residuals - Families", col = "#11A797", las = 2) # plotting the results. 
```

Figure 5 and Figure 6 show the Standardized residuals of the Chi-square test 
for order and families of class aves, respectively. 

## **Discussion**

The temporal analysis of the dataset revealed intriguing patterns in collection 
dynamics over the last three centuries, with a significant acceleration post-1870s. 
This period corresponds to heightened global exploration and scientific expeditions,
emphasizing the historical context that shaped museum collections.[@bakker2020global] 
The distribution of records across 190 nations underscores the global reach of these
collections, with concentrations in the United States and notable undersampling in 
central Africa. These patterns prompt consideration of historical, geopolitical, 
and ecological factors influencing collection efforts, offering a nuanced 
perspective on the evolution of biodiversity documentation.[@bakker2020global] 

Zooming into the taxonomic exploration of class Aves, the study reveals a 
comprehensive representation of avian biodiversity across 41 orders and 255 families. 
This rich taxonomic spectrum underscores the importance of museum collections in 
providing a detailed understanding of the diversity within avian taxa. However, 
the analysis also highlights gaps and biases in the representation of certain 
orders and families, raising questions about the factors influencing collection
priorities. A few potential reasons for the underrepresentation of certain 
taxa could be habitat inaccessibility, research focus, conservation status and 
religious believes as explained in Bakker et al., 2020 [@bakker2020global].
This exploration of taxonomic nuances enhances the depth of understanding 
regarding the completeness and comprehensiveness of the available avian specimen 
collections.  

Transitioning to tissue collections, the results shed light on the delayed 
inception in the 19th century due to challenges in reliable tissue preservation 
methods. The pronounced surge in "modern" tissue collection from the late 1970s, 
particularly in the United States, points towards technological advancements and 
shifting scientific paradigms. This shift's implications impact molecular research, 
genetic studies, and the understanding of diseases affecting avian populations.[@bi2013unlocking; @besnard2014museums] 
The global void in tissue representation, encompassing only 145 countries, 
raises questions about accessibility, infrastructure, and collaborative efforts 
in scientific research. Data shows a considerable amount of tissue repository 
pre-1950s, which is unlikely given that there were no reliable methods to preserve 
the tissue samples.[@seutin1991preservation] Hence, the records could be either 
due to human error or toe pad clippings cut much later in the collection 
assigned as a tissue. Thus, this opens the scope of the global landscape of 
scientific exploration, emphasizing the need for inclusivity and equity in the 
representation of biological diversity in tissue collections. 

In a broader context, this study contributes to the ongoing dialogue on the 
significance of museum collections as invaluable reservoirs of biodiversity 
information. It reinforces the dual role of collections as historical records 
and contemporary scientific resources, urging continued support for the digitization 
and accessibility initiatives that enhance the utility of museums. These findings 
call attention to the importance of targeted specimen and tissue collection 
initiatives to fill gaps in representation, fostering a more holistic understanding 
of avian specimen collections. Ultimately, this study could catalyze future 
discussions and actions to advance the integrity and relevance of natural history 
collections in the face of evolving scientific landscapes and global challenges. 

## **Caveats** 

As with any scientific study, this study is subject to certain constraints and 
may only comprehensively address some pertinent questions. The reliance on a 
publicly available database, limited to data until 2015, is acknowledged as a 
potential limitation. Subsequent to this period, concerted efforts have been made 
to enhance the comprehensiveness of global collection databases. A notable 
proportion of records originate from the pre-colonial era, introducing a potential 
source of imprecision in determining the precise specimen collection locations. 
Furthermore, the dynamic nature of the Tree of Life taxonomy is recognized, with 
ongoing updates and revisions. It is important to note that the lower-level 
taxonomy employed in this study needed to be standardized, introducing a potential 
source of bias in the results. Thus, the inferences drawn from this study should 
be exercised cautiously, considering these inherent limitations. 

## **Acknowledgments**

I thank Dr Brant Faircloth for helping me conceptualise the project and discuss 
the results. I would also like to thank Naman Goyal, Eamon Corbett and 
Oluwaseun Akinsulire for the meaningful discussion. I thank Dr Dijing Li for the 
opportunity to conduct this study as a part of class project. 

## **Appendix**

Other R packages used in the study are listed below
tidyverse[@tidyverse], dplyr[@dplyr], lubridate[@lubridate], stringr[@stringr], 
readr[@readr], rvest[@rvest], sf[@sf], ggplot2[@ggplot2], ggthemes[@ggthemes],
cowplot[@cowplot]

## **References**



