---
date: "2023-09-28"
output:
  pdf_document: default
  html_document: default
link-citations: yes
bibliography: references_proposal.bib
indent: true
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
knitr::write_bib(file = 'packages.bib')
```

# Where the gaps are? Analysing Vertnet database to find the global gaps in avian specimen collections.

\begin{center}
Vinay K L
\end{center}


Museum and Natural history collections are important, essential and
invaluable source of earth history's biodiversity information. The
specimens in natural history collections serve as the basis for numerous
fields of basic science and research. Natural history collections also
contain deep, taxon-specific information derived from global populations
of plants, animals, fungi, and microorganisms. [@hope2018collection;
@johnson2011climate; @card2021museum]

Some background Use of specimens - maybe Edawards paper Kevin Winker
paper Some background about Vertnet Disparity between global south
collection perhaps Usefulness of the analysis


## References

 
