---
date: "2023-09-28"
output:
  html_document: default
  pdf_document: default
link-citations: yes
bibliography: references_proposal.bib
indent: true
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
knitr::write_bib(file = 'packages.bib')
```

# Mapping the Void: An In-depth Analysis of the VertNet Database to Unearth Global Gaps in Avian Specimen Collections

```{=tex}
\begin{center}
Vinay K L
\end{center}
```
## Background

Museum and Natural history collections are important, essential and
invaluable source of earth history's biodiversity information. Museum
collections, often spanning centuries, represent a treasure trove of
biological specimens and data. The specimens in natural history
collections serve as the basis for numerous fields of basic science and
research. Natural history collections also contain deep, taxon-specific
information derived from global populations of plants, animals, fungi,
and microorganisms. [@hope2018collection; @johnson2011climate;
@card2021museum] By evaluating how species evolve over time, museum
specimens offer crucial baseline data for studies of conservation and
emerging diseases (such as
[Hantavirus](https://msb.unm.edu/divisions/mammals/research/projects/hantavirus.html)
and West Nile Virus). The importance of museum collections lies in their
ability to serve as both a historical record of life on our planet and a
contemporary resource for scientific inquiry.

What initially began as cabinets filled with mounted specimens has
transformed into vast, digitized repositories. With the advent of modern
databases and digitization efforts, these collections are now more
accessible and interconnected than ever. VertNet is one such effort to
bring together museum collections records from over 250 natural-history
collections. [@constable2010vertnet] VertNet, a comprehensive repository
of biodiversity data,serves as an invaluable resource for assessing the
state of biodiversity worldwide. The exploitation of VertNet's expansive
dataset not only facilitates a profound comprehension of the immense
diversity encompassed within avian taxa but also provides the tools to
identify important gaps in our intellectual understanding. Through the
meticulous analysis of this wealth of information, we can discern
regions characterized by a dearth of avian tissue collections, taxonomic
groups that remain underrepresented in scientific investigations, and
temporal gaps. These conclusions play a crucial part in determining how
future specimen collection initiatives will proceed.

## Data

Vertnet data base is well organised and freely available from their
cloud platform and can be downloaded/accessible from their
[website](http://www.vertnet.org/resources/datatoolscode.html) and aves
database can be downloaded from [here.](http://dx.doi.org/10.7946/P2K01C)


## Questions

VertNet has developed and deployed couple of their own pipeline and packages for data manipulation. 

1) Where are the major data gaps in terms of specimen collection and tissue sample?
2) Which are the over and under represented groups/family of birds?
3) Are the global south under-represented in the museum collection?
4) Is the long-term trends in specimen accumulation increasing or decreasing?

 
## Caveats
 
Although VertNet is a good source of database it is not updated since 2016. 

## References
